\chapter{Battle of Kircholm}
The Battle of Kircholm (Lithuanian: Salaspilio m?餜s; Polish: Bitwa pod Kircholmem; Swedish: Slaget vid Kirkholm; 27 September [O.S. 17 September] 1605) was one of the major battles in the Polish-Swedish War of 1600-1611. The battle was decided in 20 minutes by a devastating charge of Polish-Lithuanian cavalry, including the Winged Hussars.[7] The battle ended in a decisive victory of the Polish-Lithuanian forces, and is remembered as one of the greatest triumphs of Commonwealth cavalry.

\section{Background}

On 27 September 1605, the Commonwealth and Swedish forces met near the small town of Kircholm (now Salaspils in Latvia, some 18 kilometres (11 mi) south-east of Riga). The forces of Charles IX of Sweden were numerically superior and were composed of 10,868 men[1] and 11 cannons. The Swedish army included two western commanders, Frederick of L𤦂eburg and Count Joachim Frederick of Mansfeld,[1] with a few thousand German and Dutch mercenaries and even a few hundred Scots.

The Polish Crown declined to raise funds for defence and send troops, only making promises they never fulfilled.[8] The army, led by the Great Hetman of Lithuania Jan Karol Chodkiewicz, was tired and starving; however, the soldiers admired their leader.[8] He promised to pay army wages from his own fortune, resulting in an influx of recruits from Lithuania. The Polish-Lithuanian Commonwealth army under Chodkiewicz was composed of roughly 1,000 infantry and 2,600 cavalry,[1] but only five cannons. However, the Polish-Lithuanian forces were well-rested, and their cavalry consisted mostly of superbly trained Winged Hussars, or heavy cavalry armed with lances.[citation needed] In contrast, the Swedish cavalry were less-well trained, armed with pistols and carbines, on poorer horses and tired after a long night's march of over 10 km in torrential rain.[3] The Polish-Lithuanian forces had a small number of Lithuanian Tatar and Cossack mercenaries, used mostly for reconnaissance.[citation needed]

\section{Deployment}

Position of both armies prior to Polish-Lithuanian attack

The Swedish forces seem to have been deployed in a checkerboard formation, made up of the infantry regiments formed into seven or eight well-spaced independent blocks, with intersecting fields of fire. The flanks were to be covered by the Swedish and German cavalry, and the cannons were placed in front of the cavalry. Charles' force was formed into four lines on the crest of a ridge. The first line consisted of four infantry battalions, cavalry in the second line, six infantry battalions in the third line, and cavalry in the fourth line.[1] The infantry battalions formed in squares of 30 by 30, with pikemen in the center and shot on the edges, and gaps between the squares allowed passage of their cavalry.[1]

Jan Karol Chodkiewicz deployed his forces in the traditional deep Polish battle formation鍟he so-called "Old Polish Order"𨫎ith the left wing significantly stronger and commanded by Tomasz D?browa, while the right wing was composed of a smaller number of hussars under Jan Piotr Sapieha, and the center included Hetman Chodkiewicz's own company of 300 hussars led by Lt. Wincenty Woyna and a powerful formation of reiters sent by the Duke of Courland Friedrich Kettler.[9] The Polish-Lithuanian infantry, mostly armed in Hungarian hajduk-style, drew up in the center. Some 280 hussars were left as a general reserve under Teodor Lacki. 